How to select predictive models for decision-making or causal inference?

### 2023-11-19 how_to_select_causal_models.tex
ignore auxilliary files
updated cover letteer
condense the paper

%\documentclass[MYBIB,STIX1COL]{WileyNJD-v2}
\documentclass[10pt,twocolumn]{article}

@@

\title{How to select predictive models for causal inference?}
\addbibresource{biblio.bib}

@@

\onecolumn
\maketitle
\begin{abstract}
    \textbf{Background:}
    As predictive models --\emph{eg} from machine learning-- give likely
    outcomes, they may be used to reason on the effect of an intervention,
    a causal-inference task. The increasing complexity of
    health data has opened the door to a plethora of models, but also the
    Pandora box of model selection: which of these models yield the most
    valid causal estimates? Causal model selection should control both outcome errors for each
    individual, treated or not treated, whereas only one outcome is observed.
    Theoretically, simple risks used in machine learning do not control causal
    effects when treated and non-treated population differ too much. \\
    \textbf{Methods:}
    More elaborate risks build proxies of the causal error using ``nuisance''
    re-weighting to compute it on the observed data. But does computing these
    nuisance adds noise to model selection? Drawing from an extensive empirical
    study based on a simulation and three semi-simulated datasets based on real
    covariates, we compare different metrics for causal model selection.\\
    \textbf{Results:} Among all metrics the classical mean squared error (ie.
    factual $\mu\mathrm{-risk}$) is worse and re-weighting it with propensity score
    ($\mu\mathrm{-risk}_{IPW}$) does not bring much improvements. The
    $R\text{-risk}$, which includes a model of mean outcome and propensity
    scores, leads to the best performances.\\
    \textbf{Conclusions:} We outline a good causal model-selection procedure
    using the so-called $R\text{-risk}$; using flexible estimators to compute
    the nuisance models on the train set; and splitting out 10\% of the data to
    compute risks.
\end{abstract}
\keywords{Model Selection, Treatment Effect, G-formula, Observational Study, Machine Learning, }
\clearpage

@@

%\idea{There is a variety of models for g-estimation}
There is an explosion of outcome modeling or machine learning methods.
For instance, many deep-learning methods have been developed for medical
image analysis \cite{shen2017deep,monshi2020deep}.
Even outcome-modeling methods specifically designed for causal
inference are numerous: Bayesian Additive Regression Trees
\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
\cite{powers_methods_2018}, causal multivariate adaptive regression
splines \cite{powers_methods_2018}, random forests
\cite{wager_estimation_2018, athey_generalized_2019},
Meta-learners \cite{kunzel_metalearners_2019}, R-learners
\cite{nie_quasioracle_2017}, Doubly robust estimation
\cite{chernozhukov_double_2018}...
%\idea{The variety of estimators calls for model selection procedures}
The wide variety of methods raises the problem
of selecting between different estimators based on the data at hand.
%
Indeed, estimates of treatment effects can vary markedly across different
predictive models. For instance, Figure \ref{fig:acic_2016_ate_heterogeneity} shows
large variations obtained across different outcome estimators on
semi-synthetic datasets \cite{dorie_automated_2019}. Flexible models
such as random forests are doing well in most settings except
when treated and untreated populations differ noticeably, in
which case a linear model (ridge) is to be preferred.
However random forests with different hyper-parameters
(max depth= 2) yield poor estimates.
A simple rule of thumb such as preferring flexible models does not work in
general; model selection is needed.

@@

Using a predictor to reason on causal effects relies on contrasting the
prediction of the outcome for a given individual with and without the
treatment --as detailed in \autoref{sec:neyman_rubin}.
%
Given various predictors of the outcome, which one should we use?
%
Standard predictive modeling or machine-learning practice selects the
predictor that minimizes the expected error.
However, this predictor may not be the best model to reason about
causal effects of an intervention, as we illustrate below.

Figure \ref{fig:toy_example} gives a toy example: the probability $Y$ of
an undesirable
outcome (\emph{eg} death), a binary treatment $A \in \{0, 1\}$, and a covariate
$X \in \mathbb R$ summarizing the patient health status (eg. the
Charlson index \cite{charlson_new_1987}). We simulate a
treatment beneficial (decreases $Y$) for patients with high Charlson scores (bad health
status) but with little effect for patients in good
condition (low Charlson scores).

@@

We generate the response functions using random bases. Basis extension methods
are common in biostatistics, \emph{eg} functional regression with splines
\cite{howe_splines_2011, perperoglou_review_2019}. By allowing the function to
vary at specific knots, they give flexible non-linear models. We use random approximation of Radial Basis Function (RBF) kernels
\cite{rahimi_random_2008} to generate the response functions. RBF use the same
process as polynomial splines but replace polynomial by Gaussian kernels. Unlike
polynomials, Gaussian kernels have decreasing influences in the input space.
This avoids unrealistic divergences of the response surfaces at the
ends of the feature space.

The number of basis functions --\emph{ie. knots}--, controls the complexity of
the ground-truth response surfaces and treatment. We first use this process to
draw the non-treated response surface $\mu_0$ and the causal effect $\tau$. We
then draw the observations from a mixture two Gaussians, for the treated and non
treated. We vary the separation between the two Gaussians to control the
overlap between treated and non-treated populations, an important parameter
for causal inference (related to $\eta$ in Proposition
\ref{theory:prop:mu_risk_ipw_bound}). Finally, we generate observed outcomes
adding Gaussian noise. We generate 1\,000 of such datasets, with
uniformly random overlap parameters. Example shown in Figure \ref{fig:simulation_examples} and details in
\ref{apd:experiments:generation}.

@@

We test model selection on a family of candidate estimators that
approximate imperfectly the data-generating process. To build such an
estimator, we first use a RBF expansion
similar to the one used for data generation.
We choose two random knots and apply a transformation of the raw data features
with a Gaussian kernel. This step
is referred as the featurization. Then, we fit a linear regression on
these
transformed features. We consider two ways of combining these steps for outcome
mode; we use common nomenclature \cite{kunzel_metalearners_2019,shen2023rctrep} to refer
to these different meta-learners that differ on how
they model, jointly or not, the treated and the non treated:
\begin{itemize}
    \item SLearner: A single learner for both population, taking the treatment as
          a supplementary covariate.

    \item SftLearner: A single set of basis functions is sampled at random for both
          populations, leading to a given feature space used to model both the treat and
          the non treated, then two
          separate different regressors are fitted on this shared representation.
    \item TLearner: Two completely different learners for each population, hence
          separate feature representations and regressors.
\end{itemize}

@@

We also use classic benchmarks of the causal-inference
literature, semi-simulated data adding a known synthetic causal effect to real --non synthetic-- covariate:
\begin{description}
    \item[ACIC 2016] \cite{dorie_automated_2019}: The dataset is based on the
        Collaborative Perinatal Project \cite{niswander_women_1972}, a RCT
        studying infants’
        developmental disorders. The initial intervention was a child’s birth
        weight $(A = 1 \text{ if weight} < 2.5 kg)$, and outcome was the child’s
        IQ after a follow-up period. The study contained $N=4\,802$ data
        points with $D=55$ features (5 binary, 27 count data, and 23
        continuous). They simulated 77 different setups varying parameters
        for treatment and response models, overlap, and interactions between treatment and
        covariates \footnote{Original R code available at
            \url{https://github.com/vdorie/aciccomp/tree/master/2016}
            to generate 77 simulations settings.}. We used 10 different seeds for
        each setup, totaling 770 dataset instances.

@@

Drawing inspiration from the TMLE literature that uses combination of flexible
machine learning methods \cite{schuler_targeted_2017}, we use as models
for the nuisances $\check e$ (respectively $\check m$) a form of meta-learner: a
stacked estimator of ridge and boosting classifiers (respectively
regressions). We select hyper-parameters with randomized search on a validation set
$\mathcal{V}$ and keep them fix for model selection
(\ref{apd:experiments:nuisances_hp} lists hyperparameters). As extreme inverse
propensity weights induce high variance, we use clipping
\cite{swaminathan_counterfactual_2015, ionides_truncated_2008} to bound
$min(\check e, 1-\check e)$ away from 0 with a fixed $\eta=10^{-10}$, ensuring
strict overlap for numerical stability.

%\paragraph{Model selection procedure}\label{semi_simulated:selection_procedure}

%We used the same training procedure procedure as in
%\ref{paragraph:selection_procedure} (Details in supp.~mat.
%\ref{apd:selection_procedure_acic}).

@@

Figure \ref{fig:acic_2016_ate_heterogeneity} shows ATE estimations for six
different models used in g-computation estimators on the 76 configurations of
the ACIC 2016 dataset. Outcome models are fitted on half of the data and
inference is done on the other half --ie. train/test with a split ratio of 0.5.
For each configuration, and each model, this train test split was repeated ten
times, yielding non parametric variance estimates
\cite{bouthillier_accounting_2021}.

@@

    \item A basis expansion of the raw features increases the problem dimension.
          Using Radial Basis Function (RBF) Nystroem transformation \footnote{We use the
              \href{https://scikit-learn.org/stable/modules/generated/sklearn.kernel_approximation.Nystroem.html}{Sklearn
                  implementation, \cite{pedregosa_scikitlearn_2011}}}, we expand the raw
          features into a transformed space. The basis expansion samples randomly a
          small number of representers in the raw data. Then,  it computes an
          approximation of the full N-dimensional kernel with these basis components,
          yielding the transformed features $z(x)$.

          We generate the basis following the original data distribution, $\left [
                  b_1 .. b_D \right ] \sim \mathbb P(x)$, with D=2 in our simulations. Then, we
          compute an approximation of the full kernel of the data generation
          process $RBF(x, \cdot) \;  with \; x \sim \mathbb P(x)$ with these
          representers: $z(x) = [RBF_{\gamma}(x, b_d)]_{d=1..D}
              \cdot Z^T \in \mathbb{R}^D$
          with $RBF_{\gamma}$ being the Gaussian kernel $K(x, y) = exp(-\gamma
              ||x-y||^2)$ and Z the normalization constant of the kernel basis,
          computed as the root inverse of the basis kernel $Z=[K(b_i, b_j)]_{i, j
              \in {1..D}}^{-1/2}$

@@

@@

The nuisances are estimated with a stacked regressor inspired by the Super
Learner framework, \cite{laan_super_2007}). The hyper-parameters are
optimized with a random search with following search grid detailed in Table
\ref{apd:experiments:nuisances_hp_grid}. All implementations come from
\href{https://scikit-learn.org/stable/}{scikit-learn}
\cite{pedregosa_scikitlearn_2011}.

>>>>

\documentclass[10pt,twocolumn]{article}

@@

\title{How to select predictive models for decision-making or causal inference?}
\addbibresource{biblio.bib}

@@

\onecolumn
\maketitle
\begin{abstract}
    \textbf{Objective:}
    As predictive models --\emph{eg} from machine learning-- give likely
    outcomes, they may be used to reason on the effect of an intervention, a
    causal-inference task. The increasing complexity of health data has opened
    the door to a plethora of models, but also the Pandora box of model
    selection: which of these models yield the most valid causal estimates?  In
    this paper, we study model selection procedures in practical settings:
    finite samples settings and without well-specification assumption.
    \textbf{Methods:}
    Elaborate risks build proxies of the causal error using ``nuisance''
    re-weighting to compute it on the observed data. But does computing these
    nuisance adds noise to model selection? Drawing from an extensive empirical
    study based on a simulation and three semi-simulated healthcare datasets
    based on real covariates, we compare different metrics for causal model
    selection.\\
    \textbf{Results:} Among all metrics the classical mean squared error (ie.
    factual $\mu\mathrm{-risk}$) is worse and re-weighting it with propensity score
    ($\mu\mathrm{-risk}_{IPW}$) does not bring much improvements. The
    $R\text{-risk}$, which includes a model of mean outcome and propensity
    scores, leads to the best performances.\\
    \textbf{Conclusions:} We outline a good causal model-selection procedure
    using the so-called $R\text{-risk}$; using flexible estimators to compute
    the nuisance models on the train set; and splitting out 10\% of the data to
    compute risks.
\end{abstract}
\keywords{Model Selection, Treatment Effect, G-formula, Observational Study, Machine Learning, }
\clearpage

@@

%\idea{There is a variety of models for g-estimation}
There is an explosion of outcome modeling or machine learning methods.
For instance, many deep-learning methods have been developed for medical
image analysis \cite{shen2017deep,monshi2020deep}.
Even outcome-modeling methods specifically designed for causal
inference are numerous: Bayesian Additive Regression Trees
\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
\cite{powers_methods_2018}, causal multivariate adaptive regression
splines \cite{powers_methods_2018}, random forests
\cite{wager_estimation_2018, athey_generalized_2019},
Meta-learners \cite{kunzel_metalearners_2019}, R-learners
\cite{nie_quasioracle_2017}, Doubly robust estimation
\cite{chernozhukov_double_2018}...
%\idea{The variety of estimators calls for model selection procedures}
The wide variety of methods raises the problem
of selecting between different estimators based on the data at hand.
%
Indeed, estimates of treatment effects can vary markedly across different
predictive models (see example from \cite{dorie_automated_2019} developed in
Appendix \ref{apd:toy_example:acic_2016_ate_variability}).

@@

Using a predictor to reason on causal effects relies on contrasting the
prediction of the outcome for a given individual with and without the
treatment --as detailed in \autoref{sec:neyman_rubin}.
%
Given various predictors of the outcome, which one should we use?
%
Standard predictive modeling or machine-learning practice selects the predictor
that minimizes the expected error. However, this predictor may not be the best
model to reason about causal effects of an intervention as Figure
\ref{fig:toy_example} illustrates. Consider the probability $Y$ of an undesirable outcome
(\emph{eg} death), a binary treatment $A \in \{0, 1\}$, and a covariate $X \in
\mathbb R$ summarizing the patient health status (eg. the Charlson index
\cite{charlson_new_1987}). We simulate a treatment beneficial (decreases $Y$)
for patients with high Charlson scores (bad health status) but with little
effect for patients in good condition (low Charlson scores).

@@

We generate the response functions using random bases. Basis extension methods
are common in biostatistics, \emph{eg} functional regression with splines
\cite{howe_splines_2011, perperoglou_review_2019}. By allowing the function to
vary at specific knots, they give flexible non-linear models. The number of
knots controls the complexity of the function. We use random approximation of
Radial Basis Function (RBF) kernels \cite{rahimi_random_2008} to generate the
outcome and treatment functions. RBF use the same process as polynomial splines
but replace polynomial by Gaussian kernels. Unlike polynomials, Gaussian kernels
have decreasing influences in the input space. This avoids unrealistic
divergences of the response surfaces at the ends of the feature space. We
generate 1\,000 datasets based on this functions, with random overlap
parameters. Example shown in Figure \ref{fig:simulation_examples} and details in
\ref{apd:experiments:generation}.

@@

We compare different candidate estimators that approximate imperfectly the
data-generating process. To build such estimators, we first use a RBF
expansion similar to the one used for data generation. We choose two random
knots and apply a transformation of the raw data features with a Gaussian
kernel. This step is referred as the featurization. Then, we fit a linear
regression on these transformed features. We consider two ways of combining
these steps for outcome mode; we use common nomenclature
\cite{kunzel_metalearners_2019,shen2023rctrep} to refer to these different
meta-learners that differ on how they model, jointly or not, the treated and the
non treated:
\begin{itemize}
    \item SLearner: A single learner for both population, taking the treatment as
          a supplementary covariate.
    \item SftLearner: A single set of basis functions is sampled at random for both
          populations, leading to a given feature space used to model both the treat and
          the non treated, then two
          separate different regressors are fitted on this shared representation.
    \item TLearner: Two completely different learners for each population, hence
          separate feature representations and regressors.
\end{itemize}

@@

We also use three benchmarks of the causal-inference literature, semi-simulated
data adding a known synthetic causal effect to real --non synthetic-- healthcare
covariate. ACIC 2016 \cite{dorie_automated_2019} is based on the Collaborative
Perinatal Project \cite{niswander_women_1972}, a RCT studying infants’
developmental disorders containing 4,802 indivduals and 55 features. We used 770
dataset instances: 10 random seeds for each of the 77 simulated settings for the
treatment and outcomes.  ACIC 2018 \cite{shimoni_benchmarking_2018} simulated
treatment and outcomes for the Linked Births and Infant Deaths Database (LBIDD)
\cite{macdorman_infant_1998} with $D=177$ covariates. We used all 432 datasets
of size $N=5\,000$.  Twins \cite{louizos_causal_2017} is an augmentation of real
data on twin births and mortality rates \cite{almond_costs_2005}. There are
$N=11\,984$ samples, and $D=50$ covariates for which we simulated 1,000
different treatment allocations. Datasets details are given in Appendix
\ref{apd:experiments:datasets}.

@@

Drawing inspiration from the TMLE literature that uses combination of flexible
machine learning methods \cite{schuler_targeted_2017}, we use a meta-learner as models
for the nuisances $\check e$ (respectively $\check m$): a
stacked estimator of ridge and boosting classifiers (respectively
regressions). We select hyper-parameters with randomized search on a validation set
$\mathcal{V}$ and keep them fix for model selection
(\ref{apd:experiments:nuisances_hp} lists hyperparameters).

@@

Figure \ref{fig:acic_2016_ate_heterogeneity} shows ATE estimations for six
different models used in g-computation estimators on the 76 configurations of
the ACIC 2016 dataset. Outcome models are fitted on half of the data and
inference is done on the other half --ie. train/test with a split ratio of 0.5.
For each configuration, and each model, this train test split was repeated ten
times, yielding non parametric variance estimates
\cite{bouthillier_accounting_2021}. Figure \ref{fig:acic_2016_ate_heterogeneity}
shows large variations obtained across different outcome estimators on
semi-synthetic datasets \cite{dorie_automated_2019}. Flexible models such as
random forests are doing well in most settings except when treated and untreated
populations differ noticeably, in which case a linear model (ridge) is to be
preferred. However random forests with different hyper-parameters (max depth= 2)
yield poor estimates. A simple rule of thumb such as preferring flexible models
does not work in general; model selection is needed.

@@

    \item A basis expansion of the raw features increases the problem dimension.
          Using Radial Basis Function (RBF) Nystroem transformation \footnote{We
          use the
          \href{https://scikit-learn.org/stable/modules/generated/sklearn.kernel_approximation.Nystroem.html}{Sklearn
          implementation, \cite{pedregosa_scikitlearn_2011}}}, we expand the raw
          features into a transformed space. The basis expansion samples
          randomly a small number of representers in the raw data. Then,  it
          computes an approximation of the full N-dimensional kernel with these
          basis components, yielding the transformed features $z(x)$. The number
          of basis functions --\emph{ie. knots}--, controls the complexity of
          the ground-truth response surfaces and treatment. We first use this
          process to draw the non-treated response surface $\mu_0$ and the
          causal effect $\tau$. We then draw the observations from a mixture two
          Gaussians, for the treated and non treated. We vary the separation
          between the two Gaussians to control the overlap between treated and
          non-treated populations, an important parameter for causal inference
          (related to $\eta$ in Proposition
          \ref{theory:prop:mu_risk_ipw_bound}). Finally, we generate observed
          outcomes adding Gaussian noise.

          More formally, we generate the basis following the original data
                  distribution, $\left [ b_1 .. b_D \right ] \sim \mathbb P(x)$,
                  with D=2 in our simulations. Then, we compute an approximation
                  of the full kernel of the data generation process $RBF(x,
                  \cdot) \;  with \; x \sim \mathbb P(x)$ with these
                  representers: $z(x) = [RBF_{\gamma}(x, b_d)]_{d=1..D} \cdot
                  Z^T \in \mathbb{R}^D$ with $RBF_{\gamma}$ being the Gaussian
                  kernel $K(x, y) = exp(-\gamma ||x-y||^2)$ and Z the
                  normalization constant of the kernel basis, computed as the
                  root inverse of the basis kernel $Z=[K(b_i, b_j)]_{i, j \in
                  {1..D}}^{-1/2}$

@@

\subsection{ Details on the semi-simulated datasets}\label{apd:experiments:datasets}

\begin{description}
    \item[ACIC 2018] \cite{dorie_automated_2019}:  The initial intervention was a child’s birth
        weight $(A = 1 \text{ if weight} < 2.5 kg)$, and outcome was the child’s
        IQ after a follow-up period. The study contained $N=4\,802$ data
        points with $D=55$ features (5 binary, 27 count data, and 23
        continuous). They simulated 77 different setups varying parameters
        for treatment and response models, overlap, and interactions between treatment and
        covariates \footnote{Original R code available at
            \url{https://github.com/vdorie/aciccomp/tree/master/2016}
            to generate 77 simulations settings.}. We used 10 different seeds
        for each setup, totaling 770 dataset instances.

@@

The nuisances are estimated with a stacked regressor inspired by the Super
Learner framework, \cite{laan_super_2007}). The hyper-parameters are
optimized with a random search with following search grid detailed in Table
\ref{apd:experiments:nuisances_hp_grid}. All implementations come from
\href{https://scikit-learn.org/stable/}{scikit-learn}
\cite{pedregosa_scikitlearn_2011}.  As extreme inverse propensity weights induce
high variance, we use clipping \cite{swaminathan_counterfactual_2015,
ionides_truncated_2008} to bound $min(\check e, 1-\check e)$ away from 0 with a
fixed $\eta=10^{-10}$, ensuring strict overlap for numerical stability.
